Viscous Erosion in Porous Media

### 2018-04-03 docs/methods/erosion_methods.tex
NM

The boundary of each body will be represented by a grid of collocation
points. If we were to simply evolve the interfaces
under~\eqref{eqn:Vn2}, this grid would become distorted, which, if
unchecked, could lead to severe tangling and numerical instability. We
therefore introduce an artificial tangential velocity $\Vs$ that always
keeps the grid points equally spaced in arclength \cite{hou-low-she1994}. The appropriate tangential velocity satisfies 
\begin{align}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn} \, .
\end{align}
The evolution of the interface is then given by the combination of the normal and tangential velocity
\begin{align}
\dot{\xx}(t) = \Vn \nn + \Vs \ss \, .
\end{align}
Since $\Vs$ is tangential, it does not modify the physics of erosion, only the parameterization of the boundaries.

@@

First, we must address a second potential source of numerical
instability. The erosion law~\eqref{eqn:Vn} depends on the {\em absolute
value} of shear stress, $\atau$, which exhibits a corner anywhere $\tau$
changes sign (i.e.~at a stagnation point). The corner in $\atau$
introduces high frequencies into~\eqref{eqn:thetaevo1} which could result in either numerical instability or an extremely stringent step-size restriction. To avoid these difficulties, we smooth $\atau$ via a narrow Gaussian filter of width $\sigma$
\begin{align}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma}
 \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha' \, ,
\end{align}
with corresponding normal and tangential interface velocities
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align}
This smoothing further suppresses the feedback of high frequencies between $\atau$ and the body shape.

@@

To better understand the shape evolution, Fig.~\ref{shrink_intface}(a) shows several superimposed interfaces at evenly spaced times. The interfaces are color-coded by normalized time $t/t_f$, where $t_f = 1.79 \times 10^{-2}$ is the time at which the body vanishes in the simulation. As seen in Fig.~\ref{shrink_intface}(a), the body converges to a fairly slender, for-aft symmetric shape that has corners at its front and back --- somewhat reminiscent of an American football.
Fig.~\ref{shrink_intface}(b) shows how the absolute shear stress, $\atau$, is distributed along each of these interfaces. The normalized arclength $s/L$ runs counterclockwise around the body, with $s/L = $ 0 and 0.5 corresponding to the rear and front stagnation points respectively. At early times (yellow curves), the stress varies significantly, with more stress concentrated at the top and bottom of the body. However, as erosion alters the body's shape, the stress becomes more uniformly distributed over the entire surface (red curves). The absolute stress $\atau$ shows slight peaks near the stagnation points as the bodies nears the final stages of erosion (dark red). Closer inspection reveals that the stress actually vanishes {\em at} the stagnation points, but in a region that becomes increasingly narrow with time.

@@

	As the body shrinks in accordance with \eqref{area_predict}, the drag that it exerts decreases at a related rate, offering a second check on our numerics. In Fig.~\ref{area_drag}(b), we show the horizontal drag, $F_D$, on the simulated body as it changes in time (black curve). The drag is calculated using \eqref{eqn:drag} and, in the context of slender-body theory, this should be interpreted as the drag-per-unit-span. As seen in the figure, $F_D$ decreases gradually at first, then abruptly drops to zero in the final moments as the body vanishes. Though it may seem peculiar, the sudden drop is consistent with the 2D Stokes law in which drag vanishes {\em logarithmically} with body size. Additionally, \eqref{eqn:drag} permits us to calculate the separate contributions of pressure and viscous drag, which we show by the blue and green curves in Fig.~\ref{area_drag}(b) respectively. Initially, pressure and viscous drag are comparable, with the circular geometry having a slightly higher level of pressure drag. As the body erodes and assumes a more slender form, though, the pressure drag decreases more rapidly, leaving viscous drag to dominate. This behavior is associated with convergence to a drag-minimizing profile, which will be discussed in Section \ref{LimitingShape}.

@@

	We show in Fig.~\ref{fixed_intface}(a) several interfaces of this modified erosion process, with the same spatial resolution, $N_\iin = 1024$, as before. Notice the body obtains a similar profile as observed previously, but now with fixed area. Importantly, more numerical accuracy is available to quantify the terminal state. The inset in Fig.~\ref{fixed_intface}(a) zooms on the rear corner demonstrate that, due to our numerical regularization, the body is actually smooth on a fine scale. We show in Fig.~\ref{fixed_intface}(b) the stress distributions on the same interfaces. As observed previously, the stress becomes more evenly distributed as erosion reshapes the body. The slight rise in shear near the front/rear corners that was observed in Fig.~\ref{shrink_intface}(b) is no longer present, suggesting that this effect was due to loss of accuracy. Rather, the late-time stress distributions in Fig.~\ref{fixed_intface}(b) decrease monotonically to zero near the stagnation points. In this figure, time is normalized by the vanishing time $t_f = 1.79 \times 10^{-2}$ from the shrinking simulations (i.e.~Fig.~\ref{shrink_intface}). Curiously, about 3 multiples of $t_f$ are required to reach a visible steady state, which contrasts with previous studies on high-Reynolds-number erosion in which a steady state usually emerges before the body vanishes \cite{moo-ris-chi-zha-she2013}.

@@

	These fixed-area simulations enable us to precisely characterize the changing geometry as it approaches the terminal state. In particular, we measure the aspect ratio of the body (the width to height) and the opening angle of the front/rear corners. In these tests, we keep the spatial resolution fixed, $N_\iin = 1024$, and show three different cases of the smoothing parameters, $\eps$ and $\sigma$, to assess how the numerical regularization influences shape. In the three cases, we keep $\eps = \sigma$ and set them to the values 20/1024, 10/1024, and 5/1024 respectively. Figure~\ref{fig:arangle}(a) shows the aspect ratio as it varies in time. The initially circular geometry has an aspect ratio of unity, and, as erosion carves a more slender form, the aspect ratio increases monotonically. The three simulations are all very close to one another, demonstrating convergence as $\eps, \sigma \to 0$. In the simulation with these least smoothing ($\eps = \sigma = 5/1024$), the aspect ratio tends to a value of 2.7 in long time.

@@

	To more thoroughly examine the influence of the smoothing parameters, we report in Table~\ref{table:arangle} the long-time values of the aspect ratio and opening angle for each of the three simulations. Since $N_\iin = 1024$ is fixed, the values $\eps = \sigma = $ 20/1024, 10/1024, and 5/1024 roughly correspond to smoothing over the nearest 20, 10, and 5 grid points respectively. The table shows that both the aspect ratio and opening angle converge with first-order as $\eps, \sigma \to 0$. In addition, we show an estimate of the uncertainty involved in our angle measurement, obtained by varying the degree of the fitting polynomial and the size of the excluded region around the stagnation points. The estimates show that as $\eps, \sigma \to 0$ the angle-measurements become more precise and converge to the predicted value of $102^{\circ}$.

>>>>

The boundary of each body will be represented by a grid of collocation
points. If we were to simply evolve the interfaces
under~\eqref{eqn:Vn2}, this grid would become distorted, which, if
unchecked, could lead to severe tangling and numerical instability. We
therefore introduce an artificial tangential velocity $\Vs$ to keep the grid points equally spaced in arclength \cite{hou-low-she1994}. The appropriate tangential velocity satisfies 
\begin{align}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn} \, .
\end{align}
The evolution of the interface is then given by the combination of the normal and tangential velocity
\begin{align}
\dot{\xx}(t) = \Vn \nn + \Vs \ss \, .
\end{align}
Since $\Vs$ is tangential, it does not modify the physics of erosion, only the parameterization of the boundaries.

@@

First, we must address a second potential source of numerical
instability. The erosion law~\eqref{eqn:Vn} depends on the {\em absolute
value} of shear stress, $\atau$, which exhibits a corner anywhere $\tau$
changes sign, i.e.~at a stagnation point. The corner in $\atau$
introduces high frequencies into~\eqref{eqn:thetaevo1} which could result in either numerical instability or an extremely stringent step-size restriction. To avoid these difficulties, we smooth $\atau$ via a narrow Gaussian filter of width $\sigma$
\begin{align}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma}
 \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha' \, ,
\end{align}
with corresponding normal and tangential interface velocities
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align}
This smoothing further suppresses the feedback of high frequencies between $\atau$ and the body shape.

@@

To better understand the shape evolution, Fig.~\ref{shrink_intface}(a) shows several superimposed interfaces at evenly spaced times. The interfaces are color-coded by normalized time $t/t_f$, where $t_f = 1.79 \times 10^{-2}$ is the time at which the body vanishes in the simulation. As seen in Fig.~\ref{shrink_intface}(a), the body converges to a fairly slender, for-aft symmetric shape that has corners at its front and back --- somewhat reminiscent of an American football.
Fig.~\ref{shrink_intface}(b) shows how the absolute shear stress, $\atau$, is distributed along each of these interfaces. The normalized arclength $s/L$ runs counterclockwise around the body, with $s/L = $ 0 and 0.5 corresponding to the rear and front stagnation points respectively. At early times (yellow curves), the stress varies significantly, with more stress concentrated at the top and bottom of the body. However, as erosion reshapes the body, the stress becomes more uniformly distributed over the entire surface (red curves). The absolute stress $\atau$ shows slight peaks near the stagnation points as the bodies nears the final stages of erosion (dark red). Closer inspection reveals that the stress actually vanishes {\em at} the stagnation points, but in a region that becomes increasingly narrow with time.

@@

	As the body shrinks in accordance with \eqref{area_predict}, the drag that it exerts decreases at a related rate, offering a second check on our numerics. In Fig.~\ref{area_drag}(b), we show the horizontal drag, $F_D$, on the simulated body as it changes in time (black curve). The drag is calculated using \eqref{eqn:drag} and, in the context of slender-body theory, this should be interpreted as the drag-per-unit-span. As seen in the figure, $F_D$ decreases gradually at first, then abruptly drops to zero in the final moments of the body's existence. Though it may seem peculiar, the sudden drop is consistent with the 2D Stokes law in which drag vanishes {\em logarithmically} with body size. Additionally, \eqref{eqn:drag} permits us to calculate the separate contributions of pressure and viscous drag, which we show by the blue and green curves in Fig.~\ref{area_drag}(b) respectively. Initially, pressure and viscous drag are comparable, with the circular geometry having a slightly higher level of pressure drag. As the body erodes and assumes a more slender form, though, the pressure drag decreases more rapidly, leaving viscous drag to dominate. This behavior is associated with convergence to a drag-minimizing profile, which is discussed in Section \ref{LimitingShape}.

@@

	We show in Fig.~\ref{fixed_intface}(a) several interfaces of this modified erosion process, with the same spatial resolution, $N_\iin = 1024$, as before. Notice the body obtains a similar profile as observed previously, but now with fixed area. Importantly, more numerical accuracy is available to quantify the terminal state. The inset in Fig.~\ref{fixed_intface}(a) zooms on the rear corner demonstrate that, due to our numerical regularization, the body is actually smooth on a fine scale. We show in Fig.~\ref{fixed_intface}(b) the stress distributions on the same interfaces. As observed previously, the stress becomes more evenly distributed as erosion reshapes the body. The slight rise in shear near the front/rear corners that was observed in Fig.~\ref{shrink_intface}(b) is no longer present, suggesting that this effect was due to loss of accuracy. Rather, the late-time stress distributions in Fig.~\ref{fixed_intface}(b) decrease monotonically to zero near the stagnation points. In this figure, time is normalized by the vanishing time $t_f = 1.79 \times 10^{-2}$ from the shrinking simulations (i.e.~Fig.~\ref{shrink_intface}). Curiously, about 3 multiples of $t_f$ are required to reach a visible steady state, which contrasts with previous studies on high-Reynolds-number erosion in which a steady state usually emerged before the body disappeared \cite{moo-ris-chi-zha-she2013}.

@@

	These fixed-area simulations enable us to precisely characterize the changing geometry as it approaches the terminal state. In particular, we measure the aspect ratio of the body (the width to height) and the opening angle of the front/rear corners. In these tests, we keep the spatial resolution fixed, $N_\iin = 1024$, and show three different cases of the smoothing parameters, $\eps$ and $\sigma$, to assess how the numerical regularization influences shape. In the three cases, we keep $\eps = \sigma$ and set them to the values 20/1024, 10/1024, and 5/1024 respectively. Loosely, these values correspond to smoothing over the nearest 20, 10, and 5 grid points respectively. Figure~\ref{fig:arangle}(a) shows the aspect ratio as it varies in time. The initially circular geometry has an aspect ratio of unity, and, as erosion carves a more slender form, the aspect ratio increases monotonically. The three simulations are all very close to one another, demonstrating convergence as $\eps, \sigma \to 0$. In the simulation with these least smoothing ($\eps = \sigma = 5/1024$), the aspect ratio tends to a value of 2.7 in long time.

@@

	To more thoroughly examine the influence of the smoothing parameters, we report in Table~\ref{table:arangle} the long-time values of the aspect ratio and opening angle for each of the three simulations. The table shows that both the aspect ratio and opening angle converge with first-order as $\eps, \sigma \to 0$. In addition, we show an estimate of the uncertainty involved in our angle measurement, obtained by varying the degree of the fitting polynomial and the size of the excluded region around the stagnation points. The estimates show that as $\eps, \sigma \to 0$ the angle-measurements become more precise and converge to the predicted value of $102^{\circ}$.
